\section{Learning}\label{sec:learning}

Now that the structure of a \gls{nn} has been detailed, attention must be turned to how it is modified in order to accomplish the assigned task.
To do so, attention must be focused on what lies at the very core of \gls{dl}: the learning process itself.
As a reminder, the objective of a \gls{nn} is to adjust its parameters so that it becomes a good approximation of a target function.
That is, for a given architecture $\hat{f}$, the goal is to find the set of parameters $\theta$ that best approximates the target function $f$.
Mathematically, this objective is expressed as:
\[
\min_{\theta} \frac{1}{N} \sum_{i=1}^{N} \mathcal{L}(\hat{f}_{\theta}(x^{(i)}), y^{(i)}).
\]

However, the \gls{nn} $\hat{f}$ used to approximate the target function $f$ is highly complex.
It involves a large number of parameters as well as many nonlinear components.
As a result, an exact solution cannot be found by directly studying the properties of the function.
So, one must therefore rely on approximate methods to optimize \gls{nn}s.

The class of algorithms used for this optimization is known as gradient descent methods.
To develop an intuition, Figure~\ref{fig:simple_gradient_descent} illustrates gradient descent applied to the simple convex loss function $\mathcal{L}(w_1,w_2)=w_1^2+w_2^2$, where the red path shows the parameter updates converging to the minimum.

\input{figures/deepl_learning/gradient_descent/simple_grandient_descent/simple_gradient_descent.tex}

For simplicity, the focus here is on \gls{sgd} with a batch size of 1.
As in previous cases, this choice is made because this setting captures the essential intuition behind the training method.

The first step of \gls{sgd} is to randomly initialize the parameter vector $\theta \in \mathbb{R}^P$ for a given network architecture $\hat{f}$.
We denote this initial state by $\theta^{(1)}$.
The goal is then to iteratively modify this vector in order to improve the network's ability to approximate the target function.

To determine how to update the parameters, the effect of small variations in each parameter on the loss $\mathcal{L}$ is studied.
To analyze how parameter variations affect the loss, a fundamental mathematical tool known as the gradient is used.
It is defined as the collection of partial derivatives of a function with respect to its variables.
The gradient measures the effect of a small variation of each parameter on the loss.

In our setting, the objective is to understand, for a given example $(x^{(i)}, y^{(i)})$, how the parameters should be adjusted so that the \gls{nn} prediction $\hat{y}^{(i)}$ becomes closer to the true output $y^{(i)}$.
To achieve this, the partial derivatives of the loss function $\mathcal{L}$ with respect to each parameter vector $\theta$ are computed:
\[
\nabla_{\theta} \mathcal{L}(\hat{f}_{\theta}(x^{(i)}), y^{(i)}) = \left[ \frac{\partial \mathcal{L}(\hat{f}_{\theta}(x^{(i)}), y^{(i)})}{\partial w_1}, \dots, \frac{\partial \mathcal{L}(\hat{f}_{\theta}(x^{(i)}), y^{(i)})}{\partial w_P} \right].
\]

\input{figures/deepl_learning/gradient_descent/multiple_gradient_descents_non_convex/multiple_gradient_descents_non_convex.tex}

Since both the \gls{nn} and the loss function are compositions of differentiable functions, these derivatives can be efficiently computed using the chain rule.
Each component of the gradient $\frac{\partial \mathcal{L}(\dots)}{\partial w_k} \; \forall k \in [1, \dots, P]$ indicates how a small increase in the corresponding parameter $w_k$ affects the loss.
More precisely, for a given parameter, a positive value in the corresponding gradient component indicates that increasing this parameter will locally increase the loss.
Conversely, a negative value indicates that increasing it will locally decrease the loss.

Now that information is available on how changes in the parameters affect the loss, the objective is to minimize it.
To do so, the parameters are updated in the direction opposite to the gradient.
Indeed, the gradient is followed in the opposite direction because it indicates the direction in which the loss increases, whereas the objective is to reduce it.
This leads us to the gradient descent update rule:
\[
\theta^{(n+1)} = \theta^{(n)} - \eta \nabla_{\theta^{(n)}} \mathcal{L}(\hat{f}_{\theta^{(n)}}(x^{(i)}), y^{(i)}),
\]

where $\eta > 0$ is the learning rate, which controls the step size.
This parameter must remain small, as the gradient only provides reliable information in a local neighborhood of the current parameters.
As a result, a single step is not sufficient to significantly reduce the loss, so the process is iterated many times.
Producing a sequence of parameter vectors $\theta_1, \theta_2, \dots$ that progressively improve the model.
In practice, training is stopped when successive updates no longer produce a significant decrease in the loss.

It has been mathematically shown that gradient descent applied to a convex loss with respect to the parameters converges to a global optimum, independently of the initialization.
However, \gls{nn} training involves highly non-convex loss surfaces, so the optimization trajectory becomes strongly dependent on the initial conditions, as illustrated in Figure~\ref{fig:multiple_gradient_descents_non_convex}, where different starting points lead to distinct local minima on standard non-convex functions.

Despite the lack of theoretical guarantees of convergence to a global optimum in this setting, the simple procedure of random initialization followed by iterative gradient-based updates is sufficient in practice to train highly complex \glspl{nn}.
This empirical success is often attributed to the expressive power of \gls{nn}s, even though a complete theoretical understanding of this phenomenon remains an open question.