\section{Objective}\label{sec:objective}

Now that the main developments that led to the rise of \gls{dl} have been outlined, it is necessary to formalize what is actually being achieved when performing \gls{dl}.
The range of problems that can be addressed with \gls{dl} is extremely broad.
To ensure clarity, the supervised learning setting will be considered, specifically applied to a regression task.
This framework serves as a standard reference in the field, as all building blocks of \gls{dl} applications are built upon this fundamental formalism.

In this setting, learning consists of progressively shaping a \gls{nn} that serves as an approximation of a target function.
At initialization, the network does not yet provide a meaningful representation of this function.
Since a \gls{nn} is a parametric function, its parameters are progressively updated during training so as to make it a better approximation of the target function.
It gradually becomes capable of capturing the relationship between inputs and outputs through successive parameter updates during training.
For this reason, \gls{dl} is naturally situated within the framework of machine learning.
A standard definition of machine learning is:
\begin{quote}
\emph{``A computer program is said to learn from experience $E$ with respect to some class of tasks $T$ and performance measure $P$, if its performance at tasks in $T$, as measured by $P$, improves with experience $E$.''} \cite{mitchell1997machine}
\end{quote}

When this formulation is adapted to \gls{dl}, the following correspondences hold:
\begin{itemize}
\item The computer program corresponds to the learning algorithm, which is responsible for adjusting the network's parameters.
\item The task $T$ consists of learning a \gls{nn} that provides a good approximation of the target function.
\item The performance measure $P$ corresponds to a function used to evaluate the quality of this approximation.
\item The experience $E$ takes the form of a dataset composed of input--output pairs derived from the function to be approximated.
\end{itemize}

The mathematical formalism used in this chapter is now introduced in order to remove any ambiguity in the developments that follow.
The function to be approximated is denoted by $f$.
As with any function, it defines a mapping from an input to an output: $x \in \mathcal{X}$ denotes an input and $y \in \mathcal{Y}$ the corresponding output.
The function $f$ is therefore characterized by an input space $\mathcal{X}$ and an output space $\mathcal{Y}$, and can be formally written as $f : \mathcal{X} \to \mathcal{Y}$.
For simplicity of exposition and notation, both $\mathcal{X}$ and $\mathcal{Y}$ are assumed to be vector spaces throughout, while keeping in mind that this framework can be readily generalized to more settings.
Throughout this manuscript, the notation $\hat{\cdot}$ indicates that a given object is an approximation of another.
Since the goal of the trained \gls{nn} is to approximate $f$, this approximation is denoted by $\hat{f} : \mathcal{X} \to \mathcal{Y}$.
Given an input $x \in \mathcal{X}$, the output of the approximating function $\hat{f}(x)$ is denoted by $\hat{y}$.

It is important to emphasize that the target function $f$ is generally not known explicitly.
If it were, there would be no need to learn an approximation, since the true function would already provide the exact mapping.
In practice, only a finite set of observations generated by $f$ is available, organized in what is called a dataset.
To distinguish between the different samples, an index $i$ is introduced, allowing input--output pairs of the form $(x^{(i)}, y^{(i)})$ to be defined.
This leads to the following formulation for a dataset $\mathcal{D}$ of size $N$, where $N$ denotes the number of available examples:
\[
    \mathcal{D} = \{(x^{(i)}, y^{(i)})\}_{i=1}^{N}.
\]

Now that the way the target function $f$ is represented has been made explicit, one can ask what it means for a function $\hat{f}$ to be a good approximation of this function.
It means that, for a given input $x^{(i)}$, the \gls{nn} output $\hat{y}^{(i)}$ is close to the true output $y^{(i)}$.
To quantify the quality of this approximation, a function $\mathcal{L}$ is introduced that measures the discrepancy between predictions and target values.
In \gls{dl}, this function called the loss function plays a central role as it guides the learning process.
In regression settings, a common choice is the \gls{mse}, defined as:
\[
    \mathcal{L}(\hat{y}, y) = (\hat{y} - y)^2.
\]
Thanks to this loss, it is now possible to rigorously define the performance measure $P$.
This performance measure corresponds to the average loss over the dataset and can therefore be written as:
\[
    P = \frac{1}{N} \sum_{i=1}^{N} \mathcal{L}(\hat{f}(x^{(i)}), y^{(i)}).
\]

This formulation captures the mathematical core of the learning objective.
The goal is to find a \gls{nn} $\hat{f}$ that minimizes the prediction error over the dataset $\mathcal{D}$.
As can be seen, the setting considered here is very general, where the target function $f$ may take many different forms.
Consequently, the network $\hat{f}$ must be expressive enough to approximate a wide variety of such functions.
This requirement motivates the choice of \gls{nn} architectures, which are specifically designed to represent highly diverse functional mappings.