\chapter{Explainability in Reinforcement Learning}
\label{chap:explainability_reinforcement_learning}

Today, the majority of artificial intelligence agents with which humans interact are deployed in relatively controlled and low-risk environments.
These agents are currently mainly used in simulated, digital, or non-critical impact settings, such as games, recommendation systems, online advertising, or computer system optimization.
For example, agents capable of playing games such as Go operate in settings where potential errors remain limited in terms of consequences.
Even more autonomous systems, such as agents based on large language models, remain strongly constrained by human supervision mechanisms and are often restricted to tasks such as code generation or information retrieval.
Safety-critical applications integrating machine learning components are still relatively rare.

However, many potential high-impact applications such as drone control, autonomous delivery, autonomous vehicle driving, adaptive urban traffic management, optimization of smart power grids, or control of industrial robotic systems in real-world environments would require significantly higher levels of autonomy.
Introducing autonomous agents in such environments, requires the establishment of strong guarantees.
These include in particular:
\begin{itemize}
    \item validating the system’s behavior before deployment,
    \item being able to identify the origin of a malfunction in case of failure,
    \item and being able to certify that an observed error will not occur again.
\end{itemize}
Without satisfying these criteria, deploying autonomous agents in the real world remains hardly feasible.

The main challenge lies in the fact that the most performant agents currently available rely on reinforcement learning methods.
This learning process is carried out without explicit human supervision, and these agents are not able to provide explanations for the mechanisms that lead to their decisions.
This lack of explainability makes their use difficult in safety-critical contexts.
To address this limitation, a research field has emerged: Explainable Reinforcement Learning (XRL), a subfield of Explainable Artificial Intelligence (XAI).

Although relatively recent, this field is rapidly growing and offers a wide variety of approaches.
This chapter aims to present the main existing XRL techniques.
To this end, we first define what is meant by an “explanation” in the context of reinforcement learning.
We then propose a structuring of the XRL research field.
Finally, we review the state of the art following this organization.

\input{chapters/explainability_reinforcement_learning/explanation.tex}
% \input{chapters/explainability_reinforcement_learning/classification_method.tex}