\section{Conclusion}

\input{figures/latent_space/latent_space_overview.tex}

The four families of methods reviewed in this chapter, namely probing, interpretable directions, sparse autoencoders, and clustering, are summarized in Figure~\ref{fig:latent_space_overview} and all build upon the concept hypothesis introduced in Section~\ref{sec:concept_hypothesis}.
Each of them approaches the question from a different angle, but they share a common objective: to uncover how latent representations are structured and what semantic information can be extracted from them.

However, a fundamental difficulty remains.
There is currently no consensus on how to rigorously evaluate to what extent these methods genuinely provide access to the internal concepts of a model.
Most existing metrics assess reconstruction quality or linear separability, but these do not directly measure whether the discovered structures correspond to meaningful abstractions from the model's perspective.
This leaves open the question of what it means, formally, for a concept to be represented in a neural network.

Addressing this question may require contributions from disciplines beyond computer science and mathematics.
As the notion of semantics is inherently tied to human interpretation, a purely formal or computational account may be insufficient.
Insights from philosophy, cognitive science, or psychology could prove valuable in grounding the notion of concept in a way that is both theoretically rigorous and empirically tractable.
Despite these open questions, the methods studied in this chapter provide a practical foundation for analyzing the internal representations of neural networks.