\section{Concept Hypothesis}
\label{sec:concept_hypothesis}

\input{figures/latent_space/latent_space_embeddings}

In this section, we introduce the main working hypothesis underlying much of the literature on \gls{ls} analysis.  
Although rarely stated explicitly, this hypothesis is implicitly present in the recurring use of notions such as abstraction, representation, \gls{ls}, concept, and semantics.   
The goal of this section is to make this conceptual framework explicit by defining its main components.  

This hypothesis originates from the need to explain the generalization ability of deep \gls{nn} observed after training, as discussed in Section~\ref{sec:evaluation}.
A model is said to generalize when, for an input $x \in \mathcal{X}$ not seen during training, it still produces an output $\hat{y}$ close to the true output $y$.  
Such behavior cannot be explained by simple memorization of training samples.  
Instead, the model must extract and exploit underlying regularities in the target function.  
This requires a process of abstraction, as defined in Definition~\ref{def:abstraction}, whereby raw inputs are transformed into internal structures that are meaningful for the task \cite{bengio2014representationlearningreviewnew}.  

\begin{definition}{Abstraction}{abstraction}
Abstraction is the process by which a model transforms an input into a task-relevant internal form.  
It captures and restructures information in order to represent relationships that are useful for prediction or decision-making.  
\end{definition}

The ability of \gls{nn} to perform abstraction arises from their intermediate activations.
Indeed, a \gls{nn} does not directly map $x$ to $y$.  
Instead, as illustrated in Figure~\ref{fig:general_architecture_multilayer_perceptron}, it computes a sequence of intermediate activations across hidden layers.  
These activations correspond to internal encodings of the input, which we refer to as representations in the sense of Definition~\ref{def:representation}.

\begin{definition}{Representation}{representation}
A representation is the internal encoding of an input produced by a trained \gls{nn}.  
It corresponds to the activations of one or several layers and provides a transformed view of the input.  
\end{definition}

Each hidden layer refines the representation of the previous layer.  
Each transformation progressively reshapes the representation into a form more directly aligned with the output space \cite{rumelhart1986learning}.  
This process can be described as a sequence of transformations where each $h^{(l)}$ is a representation:  
\[
x \rightarrow h^{(1)} \rightarrow h^{(2)} \rightarrow \dots \rightarrow h^{(L-1)} \rightarrow \hat{y}
\]

The set of all such representations produced by a layer or a model over a dataset defines its \gls{ls}, as defined in Definition~\ref{def:latent_space}.

\begin{definition}{\gls{ls}}{latent_space}
The \gls{ls} is the set of all representations produced by a layer over a dataset.  
It is shaped by learning and organizes representations according to task-relevant properties.  
\end{definition}

Under the concept hypothesis, the analysis of the \gls{ls} is assumed to provide access to higher-level abstract structures learned by the model, as illustrated by \ref{fig:latent_space}.
These structures are referred to as concepts, understood here in a philosophically pragmatic sense.
In this framing, concepts are not evaluated as true or false ideal entities, but rather as abstract tools whose validity depends on their usefulness for a given task \cite{NEWELL198287}.
Our definition is given in Definition~\ref{def:concept}.

\begin{definition}{Concept}{concept}
A concept is a task-dependent abstraction that captures regularities in the input space relevant for predicting the target output.
\end{definition} 

This perspective is particularly well aligned with \gls{ls} analysis, where the objective is to uncover organizing principles that support the model’s behavior.
From this perspective, a given representation can be interpreted as a composition of multiple such concepts, each contributing to the final prediction.

Another property that can be considered when analyzing concepts is their relationship to human interpretation.  
This relationship is captured by the notion of semantics \cite{marconato2023interpretabilitymindbeholdercausal}, defined in Definition~\ref{def:semantics}.  
Under this definition, concepts may exhibit varying degrees of semantic content.  
Some concepts can be readily associated with human-interpretable notions, whereas others may contribute to task performance while remaining difficult to interpret.

\begin{definition}{Semantics}{semantics}
Semantics refers to the alignment between concepts in the model \gls{ls} and concepts used by a human observer.  
A structure is semantic if it can be consistently associated with a human-interpretable concept.  
\end{definition}

It is important to note that this conceptual framework is not supported by a complete theoretical understanding of deep \gls{nn}.  
Current mathematical tools do not provide a rigorous characterization of their internal representations and do not formally prove that concepts emerge as identifiable structures within \gls{ls}.
Nevertheless, this perspective constitutes a widely adopted working hypothesis in the field of representation and \gls{ls} analysis.    

The remainder of this chapter is devoted to reviewing methods for analyzing \gls{ls} and extracting the concepts they encode. We begin with probing, one of the most direct approaches for assessing whether concepts are represented in the \gls{ls}.