\section{Historical Developments}
\label{sec:rl_historical_developments}

\gls{rl} is today presented as a major field within artificial intelligence.  
However, it originates from several research traditions that progressively converged toward a common framework.  
In this section, we outline the main stages in the emergence of this field, highlighting the key ideas and milestones that shaped its evolution.

\subsection{Behavioral foundations (1900--1950)}

The origins of \gls{rl} can be traced back to 20th-century animal psychology studies \cite{thorndike1911law}.
These studies aimed to understand how behaviors are formed and modified.
The central idea emerging from this work is that learning is driven by the consequences of actions.
Behaviors followed by positive outcomes are more likely to be repeated, whereas those associated with negative outcomes tend to gradually decrease in frequency.

This perspective was later developed within behaviorism.  
In this school of psychology, rewards and punishments are used to gradually shape behavior through repeated interaction with the environment \cite{skinner2019behavior}.  
The term “reinforcement” in the context of learning comes from this school of thought, where it refers precisely to the process of strengthening a behavior through its consequences.

This idea led to the suggestion that learning principles observed in living organisms could be transferred to artificial systems.
Machines were therefore envisioned as potentially learning through reward-like mechanisms, guided by evaluative feedback signals resembling a “pleasure–pain system” \cite{turing2004intelligent}.  
Although these studies did not involve formal mathematical models, they introduced the central idea of \gls{rl}, trial-and-error learning guided by environmental feedback.

\subsection{Sequential decision-making (1950--1970)}

A second foundation of \gls{rl} comes from optimal control theory.
This field studies how to determine sequences of decisions that optimize a cumulative performance criterion over time.
In this context, the decision-making entity is called an agent.  
The agent interacts with the environment by selecting actions.  
The function that maps an observed state to the chosen action is called a policy.
The objective is to find an optimal policy, in the sense that it maximizes the expected cumulative reward over time.

This framework led to the formalization of \gls{mdp}, which provide a mathematical model for sequential decision-making \cite{howard1960mdp}.
\gls{mdp} describe a decision-maker interacting with an environment through a set of states, actions, rewards, and transition rules.
The earliest method proposed to solve \gls{mdp} is dynamic programming \cite{bellman1957dynamic}.
It proceeds by breaking down a complex decision problem into smaller, more manageable subproblems, solving each one, and combining the results to obtain an overall solution.
However, dynamic programming requires a full and exact description of the environment, which is only feasible for small problems.

\subsection{Emergence of reinforcement learning (1970--2000)}

In order to address this issue, temporal-difference learning methods were introduced \cite{sutton1988td}.  
This mechanism enables incremental learning directly from experience without requiring a complete model of the environment.  
It can be interpreted as a sample-based approximation of the Bellman equations.  

Building on these ideas, the notion of value functions emerged as a way to estimate the long-term benefit of being in a given state or taking a given action.
These functions assign a numerical score to states or state-action pairs, reflecting the expected cumulative reward an agent can obtain from that point onward.
They thus provide a compact way to evaluate and compare different courses of action.
This line of work culminated in Q-learning, which showed that optimal action-value functions can be learned directly from interaction with the environment, without requiring a model of its dynamics \cite{watkins1989qlearning,watkins1992qlearning}. 

Shortly before the turn of the millennium, the publication of the foundational reference work \cite{suttonbarto2018} contributed to the formalization of \gls{rl} as a coherent field.
From that point onward, the field can be considered well established and widely recognized as a distinct area of research.

\subsection{Consolidation of reinforcement learning (2000--2013)}

A limitation of early algorithms such as Q-learning is their reliance on explicit state–action tables. 
Although these methods remove the need to know the full dynamics of the \gls{mdp}, they still require storing a value for each possible state–action pair. 
As a result, the agent must effectively visit and maintain estimates for a potentially enormous number of states in the environment. 
This requirement becomes infeasible in large-scale \gls{mdp}, where the state space is too large to be exhaustively explored or stored in memory.

To address this issue, research shifted toward more general representations of value functions and policies. 
Instead of maintaining explicit tables, function approximation methods were introduced.  
These allowed value functions and policies to be represented using parametric models such as linear approximators or perceptron-like architectures \cite{BartoSuttonAnderson1983, tsitsiklis1996analysis}.
This change enabled generalization across states, reducing the dependence on exhaustive exploration of the state space.
The introduction of Deep Q-Networks (DQN), which demonstrated that deep \gls{nn} could be used to approximate value functions at scale \cite{mnih2013playingatarideepreinforcement}.

\subsection{Deep reinforcement learning (2013--2026)}

The introduction of \gls{dl} into \gls{rl} marked a major turning point in the field.  
A series of algorithms significantly improved scalability.
Deep Deterministic Policy Gradient extended actor–critic methods to continuous action spaces \cite{lillicrap2019continuouscontroldeepreinforcement}.  
\gls{ppo} later became a standard algorithm in \gls{rl} due to its simplicity, stability, and ability to perform in both continuous and discrete action spaces \cite{schulman2017proximalpolicyoptimizationalgorithms}.

These advances established a practical foundation for large-scale policy optimization.
Notably, AlphaZero demonstrated that the combination of deep \gls{nn}, \gls{rl}, and tree search could reach superhuman performance in the game of Go \cite{silver2017masteringchessshogiselfplay}.  
This success was later extended to even more complex real-time strategy environments such as StarCraft II, where agents must handle long-term planning, partial observability, and large action spaces \cite{samvelyan2019starcraftmultiagentchallenge}.  
Similarly, in Dota 2, large-scale multi-agent \gls{rl} systems such as OpenAI Five showed that self-play combined with deep policy optimization can achieve professional-level performance in highly dynamic and stochastic environments \cite{openai2019dota2largescale}.

More recently, \gls{rl} has expanded beyond its traditional application domains. 
With the emergence of increasingly agentic views of \gls{llm}, it has become possible to train these models to perform specific tasks using \gls{rl}-based techniques.  
In particular, \gls{rl} from human feedback has become a method for aligning \gls{llm} with human preferences.  
It does so by formalizing the problem as an \gls{mdp} and using \gls{rl} to solve it \cite{ouyang2022traininglanguagemodelsfollow}.
Similarly, DeepSeekMath shows how \gls{rl} can enhance the mathematical reasoning abilities of \gls{llm} \cite{shao2024deepseekmathpushinglimitsmathematical}.
It demonstrates that an artificial agent can learn to reason in natural language in order to accomplish a given task.
The skills acquired during this learning process also transfer to other similar tasks, leading to improved overall performance.
For instance, training a \gls{llm} with \gls{rl} on mathematical problem-solving can also enhance its programming capabilities.

Originally inspired by biological learning processes, \gls{rl} is now closer than ever to this initial intuition.
Agents learn through interaction, adapt via trial and error, and progressively improve their behavior based on experience.
This behavior is governed by deep neural architectures, which are the closest computational analog to organic brains.
More than ever, \gls{rl} stands as one of the core paradigms for achieving artificial general intelligence \cite{silver2021reward}.