\section{Objective}
\label{sec:rl_objective}

Now that the main developments shaping \gls{rl} have been outlined, the formal definition of its objective can be introduced.
In essence, \gls{rl} aims to build a system that makes sequential decisions in order to maximize a cumulative reward signal over the long term.
The purpose of this section is to translate this intuitive goal into a rigorous mathematical framework.
To this end, the presentation proceeds step by step:
\begin{enumerate}
    \item Problems are modeled using Markov Decision Processes in Subsection~\ref{subsec:markov_decision_processes}.
    \item The notion of policy, which formalizes the agent's behavior, is introduced in Subsection~\ref{subsec:policy}.
    \item The interaction between the agent and the environment is characterized through trajectories in Subsection~\ref{subsec:trajectory}.
    \item Policies are evaluated using value functions in Subsection~\ref{subsec:value_functions}.
    \item The optimal policy is defined in Subsection~\ref{subsec:optimal_policy}.
    \end{enumerate}

\subsection{Markov Decision Processes}
\label{subsec:markov_decision_processes}

The RL framework is grounded in the agent paradigm: the decision-making entity is called the agent.  
The agent perceives its environment, processes the information it receives, and acts upon the environment to influence future outcomes.  
To specify the task and the agent's capacity to perceive and act, the problem is modeled as a \gls{mdp}.

Before providing a more detailed definition, it is useful to clarify the scope of the \gls{mdp} problems considered in this manuscript.
The goal of this subsection is to introduce the notations and formalisms as they are used in modern RL practice.  
Other formulations of \gls{mdp} exist that are better suited, for instance, to proving convergence guarantees.  
Our focus here is to represent \gls{rl} as it is applied, even if this means sacrificing some mathematical proof capability.  
We therefore restrict the analysis to the following configuration:

\begin{itemize}
    \item The single-agent learning setting is considered.
    If any other agents are present in the environment, they are treated as part of the environment dynamics.

    \item Finite-horizon \gls{mdp} are considered.  
    Each interaction sequence between the \gls{mdp} and the policy has a defined beginning and a defined end.  
    Between these two points, the number of interaction steps may vary depending on the situation, but it always remains finite.

    \item The partially observable framework is adopted, where the agent does not have direct access to the true state of the system, but instead relies on limited observations that provide only partial information about the environment.  

\end{itemize}


The resulting framework can be formally described as a single-agent finite-horizon partially observable markov decision process.  
For simplicity, we will refer to this setting as an “\gls{mdp}” throughout this manuscript.  
A \gls{mdp} is defined by the tuple \((\mathcal{S}, \mathcal{O}, \mathcal{A}, \mathcal{R}, p)\):  
\begin{itemize}
    \item \textbf{\(\mathcal{S}\)}: The set of states, representing all possible configurations of the environment.  
    A state belonging to this set is denoted by \(s \in \mathcal{S}\).  
    Among these, two special subsets are distinguished: initial states, from which an interaction can begin, and terminal states, in which the interaction ends.  
    The set of initial states is denoted by \(\mathcal{S}^i\), with an initial state written as \(s^i \in \mathcal{S}^i\).  
    The set of terminal states is denoted by \(\mathcal{S}^t\), with a terminal state written as \(s^t \in \mathcal{S}^t\).
    
    \item \textbf{\(\mathcal{O}\)}: The set of observations that the agent can perceive from the environment.
    A observation belonging to this set is denoted by \(o \in \mathcal{O}\). 
    Unlike the true state \(s \in \mathcal{S}\), which is hidden from the agent, the observation \(o \in \mathcal{O}\) provides partial information about the current state.

    \item \textbf{\(\mathcal{A}\)}: The set of actions that the agent may execute.
    A action belonging to this set is denoted by \(a \in \mathcal{A}\). 
    Actions influence the dynamics of the \gls{mdp}, thereby shaping its future course.  
    They represent the agent's capacity to act upon its environment.

    \item \textbf{\(\mathcal{R}\)}: The set of rewards that the agent can receive from the environment.  
    A reward belonging to this set is denoted by \(r \in \mathcal{R}\). 
    \(\mathcal{R}\) is a subset of \(\mathbb{R}\).  
    The agent's goal is to maximize the cumulative sum of these rewards over time.

    \item \textbf{\(p\)}: The transition function, which defines the dynamics of the environment.  
    Given a state \(s\) and an action \(a\), this function defines a probability distribution over the joint outcome consisting of the next state \(s'\), the observation \(o\), and the reward \(r\).  
    This function is denoted by \(p\), and is defined as:
    \(
    p : \mathcal{S} \times \mathcal{A} \to \mathcal{P}(\mathcal{S} \times \mathcal{O} \times \mathcal{R}).
    \)
    The notation \((s', o, r) \sim p(s, a)\) means that the tuple \((s', o, r)\) is sampled from the distribution induced by \(p(\cdot \mid s, a)\).  
    The notation \(p(s', o, r \mid s, a)\) denotes the probability of observing the tuple \((s', o, r)\) under the distribution \(p(\cdot \mid s, a)\).
\end{itemize}

\subsection{Policy}
\label{subsec:policy}

The agent’s decision-making capability is modeled by a policy function.  
The policy defines the stochastic mapping between observations \(o \in \mathcal{O}\) and actions \(o \in \mathcal{A}\).  
A policy function is denoted by \(\pi\), and is defined as:
\(
\pi : \mathcal{O} \to \mathcal{P}(\mathcal{A}).
\)
The notation \(a \sim \pi(\cdot \mid o)\) means that the action \(a\) is sampled from the distribution induced by \(\pi(\cdot \mid o)\).  
The notation \(\pi(a \mid o)\) denotes the probability of selecting action \(a\) under the distribution \(\pi(\cdot \mid o)\).

\subsection{Trajectory}
\label{subsec:trajectory}

Now that both the \gls{mdp} \((\mathcal{S}, \mathcal{O}, \mathcal{A}, \mathcal{R}, p)\) and the policy \(\pi\) have been defined, the interaction between them can be characterized.  
A sequence of consecutive interactions between an environment and a policy is called a trajectory.  
A trajectory is generated through the repeated interaction between the policy and the environment, following the sequence below:
\begin{enumerate}
    \item The environment, being in state \(s\), provides an observation \(o\) to the policy \(\pi\).
    Based on this observation, the policy outputs a probability distribution over actions \(\pi(\cdot \mid o)\), from which an action is sampled:
    \(
    a \sim \pi(\cdot \mid o).
    \)

    \item The sampled action is executed in the environment. The environment then generates a transition:
    \(
    (s', o', r) \sim p(\cdot \mid s, a).
    \)
    It moves to a new state \(s'\), and returns a new observation \(o'\) and a reward \(r\).
\end{enumerate}

A trajectory \(t\) of length \(T\) can be written as:
\[
t = (s_1, o_1, a_1, r_1, s_2, o_2, a_2, r_2, \ldots, s_T, o_T, a_T, r_T).
\]

It is important to clearly distinguish between a trajectory distribution and its realizations.  
The distribution over trajectories is denoted by \(\tau\), and a particular trajectory is denoted by \(t\).
The distribution \(\tau\) is determined by the initial state \(s\), the initial observation \(o\), the environment dynamics \(p\), and the policy \(\pi\), and is denoted by \(\tau(s, o, p, \pi)\).
The notation \(t \sim \tau(s, o, p, \pi)\) means that the trajectory \(t\) is sampled from the distribution induced by \(s\), \(o\), \(p\), and \(\pi\).
The notation \(\tau(t \mid s, o, p, \pi)\) denotes the probability of observing the trajectorie \(t\) under the distribution \(\tau(\cdot \mid s, o, p, \pi)\)

An important property of the trajectory distribution is its recursive structure.
Since both the policy \(\pi\) and the transition function \(p\) are known, the distribution over trajectories starting from \(s_1\) with observation \(o_1\) can be expressed in terms of the distribution over trajectories starting from the next state \(s_2\) with observation \(o_2\).
Specifically, the probability of a trajectory \(t\) factorizes as:
\[
\tau(t \mid s_1, o_1, p, \pi) = \pi(a_1 \mid o_1) \; p(s_2, o_2, r_1 \mid s_1, a_1) \; \tau(t_{2:} \mid s_2, o_2, p, \pi),
\]
where \(t_{2:} = (s_2, o_2, a_2, r_2, \ldots, s_T, o_T, a_T, r_T)\) denotes the trajectory starting from the second step.

There exist two special types of trajectories.
First, a trajectory that terminates in a terminal state is called a terminal trajectory, and is denoted by \(t_t \sim \tau_t(s, o, p, \pi)\).
Second, a terminal trajectory that starts from an initial state and an initial observation is called an episode, and is denoted by \(t_e \sim \tau_t(s^i, o, p, \pi)\).

\subsection{Value Functions}
\label{subsec:value_functions}

To evaluate the ability of a policy \(\pi\) to collect rewards, the notion of return is introduced.
The return of a trajectory \(g(t, \gamma)\), is defined as the discounted sum of rewards collected along the trajectory:
\[
g(t, \gamma) = r_1 + \gamma r_2 + \gamma^2 r_3 + \cdots + \gamma^{T-1} r_T = \sum_{i=1}^{T} \gamma^{i-1} r_i,
\]
where \(\gamma \in [0, 1]\) is the discount factor.  
This parameter controls the time horizon of the agent's objective.  
When \(\gamma\) is close to \(0\), the agent focuses almost exclusively on immediate rewards, leading to short-sighted behavior.  
As \(\gamma\) increases toward \(1\), future rewards gain more weight, and the agent adopts a more far-sighted strategy.  
In theory, since the goal is to maximize long-term cumulative reward, the discount factor would ideally be set to \(\gamma = 1\).
In practice, however, \(\gamma\) is often chosen slightly below \(1\) (e.g., \(0.99\)) to stabilize learning, and to reflect the fact that distant future rewards are inherently more uncertain.
Throughout this manuscript, unless explicitly stated otherwise, the discount factor is assumed to be \(\gamma = 1\) to simplify notation.

The return also admits a recursive form, which is central to the dynamic programming and \gls{rl} algorithms presented later:
\[
g(t, \gamma) = r_1 + \gamma \, g(t_{2:}),
\]

From the notion of return, functions that estimate the expected return of a policy from a given starting point can be defined.  
These functions are called value functions.
The first one considered in detail is the state-value function \(V\).  
It tells us how desirable a particular state is under a given policy \(\pi\), given an observation \(o\) and environment dynamics \(p\).  
Strictly speaking, this function should be written as \(V(s, o, p, \pi)\) to reflect all its dependencies.  
However, since the environment dynamics \(p\) are fixed throughout learning, they are omitted from the notation.
To emphasize that the value is tied to a specific policy, \(\pi\) is placed as a subscript. 
This yields the compact notation \(V_\pi : \mathcal{S} \times \mathcal{O} \to \mathbb{R}\).

If \(V_\pi(s, o) > V_\pi(s', o')\), then, under policy \(\pi\), the agent prefers being in state \(s\) with observation \(o\) over state \(s'\) with observation \(o'\).  
Formally, the state-value function is defined as:
\[
V_\pi(s, o) = \mathbb{E}\Big[g(t)\Big], \quad \text{with } t \sim \tau_t(s, o, p, \pi).
\]

It is worth noting that the state-value function \(V_\pi\) is recursive, a property that follows directly from the recursive form of the return:
\[
\begin{aligned}
V_\pi(s, o) 
&= \mathbb{E}\Big[g(t)\Big], \quad \text{with } t \sim \tau_t(s, o, p, \pi) \\[4pt]
&= \mathbb{E}\Big[r_1 + \gamma \, g(t_{2:})\Big] \\[4pt]
&= \mathbb{E}\Big[r_1 + \gamma \, V_\pi(s_2)\Big],
\end{aligned}
\]

If the expectation is expressed explicitly in terms of the distributions \(p\) and \(\pi\), the Bellman equation for \(V_\pi\) is obtained:

\[
\begin{aligned}
V_\pi(s, o) 
&= \mathbb{E}\Big[r_1 + \gamma \, V_\pi(s_2)\Big], \quad \text{with } t \sim \tau_t(s, o, p, \pi) \\[4pt]
&= \sum_{a \in \mathcal{A}} \pi(a \mid o) 
   \sum_{\substack{s' \in \mathcal{S} \\ o' \in \mathcal{O} \\ r \in \mathcal{R}}} 
   p(s', o', r \mid s, a) \Big[\, r + \gamma \, V_\pi(s', o') \,\Big].
\end{aligned}
\]

Now that the value of a state has been expressed, the next step is to consider the value of taking a specific action in a given state.
Indeed, it is useful to estimate the expected future gain of an action \(a\) in state \(s\), in order to compare different actions and determine which one is the most promising under a given policy \(\pi\).
This function is called the state-action value function \(Q_\pi : \mathcal{S} \times \mathcal{A} \to \mathbb{R}\).
If \(Q_\pi(s, a) > Q_\pi(s, a')\), then, under policy \(\pi\), choosing action \(a\) in state \(s\) yields a higher expected return than choosing action \(a'\).
It is defined as:
\[
Q_\pi(s, a) = \mathbb{E}\Big[\, r + \gamma \, g(t) \,\Big], \quad \text{with } t \sim \tau_t(s', o, p, \pi), \quad (s', o, r) \sim p(\cdot \mid s, a).
\]

The function \(Q_\pi\) can be related to \(V_\pi\) by recognizing that, after taking action \(a\) in state \(s\), the remaining return from the next state \(s'\) is precisely \(V_\pi(s')\).
This yields:
\[
\begin{aligned}
Q_\pi(s, a) 
&= \mathbb{E}\Big[\, r + \gamma \, g(t) \,\Big], \quad \text{with } t \sim \tau_t(s', o, p, \pi), \quad (s', o, r) \sim p(\cdot \mid s, a) \\[4pt]
&= \mathbb{E}\Big[\, r + \gamma \, V_\pi(s', o) \,\Big] \\[4pt]
&= \sum_{\substack{s' \in \mathcal{S} \\ o' \in \mathcal{O} \\ r \in \mathcal{R}}} 
   p(s', o', r \mid s, a) \Big[\, r + \gamma \, V_\pi(s', o) \,\Big] \\[4pt]
\end{aligned}
\]

\subsection{Optimal Policy}
\label{subsec:optimal_policy}

The goal of \gls{rl} is to find a policy that maximizes, for every observation, the expected cumulative reward until a terminal state is reached.  
This policy is called the optimal policy and is denoted by \(\pi^*\).  
Formally, \(\pi^*\) is a deterministic policy that always selects the action with the highest state-action value:
\[
\pi^*(a \mid o) =
\begin{cases}
1 & \text{if } a = \underset{a' \in \mathcal{A}}{\arg\max}\; Q_{\pi^*}(s, a'), \\[6pt]
0 & \text{otherwise},
\end{cases}
\]
where \(s\) is the true state underlying the observation \(o\).

It is worth noting that defining the optimal policy in a partially observable setting is more delicate than in the fully observable case.  
Since the policy only has access to observations \(o\) rather than the true state \(s\), the observation must carry enough information to distinguish between states that require different optimal actions.  
For the sake of simplicity, we assume throughout this manuscript that this condition holds.  
This assumption is of interest when one seeks to establish mathematical proofs, but in real-world implementations it matters little.

As just shown, both \(V_\pi\) and \(Q_\pi\) can be expressed explicitly in terms of the environment dynamics \(p\) and the policy \(\pi\).
If \(p\) and \(\pi\) are fully known, one can theoretically compute the optimal policy by unfolding the graph of all possible trajectories starting from a given state \(s\) and observation \(o\).  
This is possible precisely because \(V_\pi\) and \(Q_\pi\) are recursive functions.  
Solving an \gls{mdp} in this way is known as dynamic programming.

In practice, most environments of have too many states to allow an explicit representation of the dynamics \(p\).  
Storing and computing over the full transition function becomes intractable.
To overcome this limitation, instead of directly constructing the optimal policy, an iterative approach is adopted: starting from an initial policy, it is progressively improved so that it converges toward \(\pi^*\).
This iterative improvement through interaction with the environment is precisely what constitutes learning in \gls{rl}, and it is the subject of the next section.